---
title: "individual1"
author: "Lisa McDaniel"
date: "9/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readxl)
library(tidyverse)
dat <- readxl::read_xls('../ds202lab3/GSS.xls', sheet="Data")
names(dat)
str(dat)
dat
```
```{r}
dat <- dat %>%
  rename(Sign = "Respondents astrological sign" ) %>%
  rename(Vocab = "Number words correct in vocabulary test") %>%
  rename(Origin = "Country of family origin") %>%
  rename(MarHap = "Married people happier than unmarried") %>%
  rename(Happy = "General happiness") %>%
  rename(PolParty = "Political party affiliation") %>%
  rename(Gender = "Respondents sex") %>%
  rename(School = "Highest year of school completed") %>%
  rename(Occ80 = "Rs occupational prestige score  (1980)") %>%
  rename(Occ70 = "Rs occupational prestige score  (1970)") %>%
  rename(IDnum = "Respondent id number") %>%
  rename(Ballot = "Ballot used for interview") %>%
  rename(Year = "Gss year for this respondent") %>%
  rename(Married = "Marital status")

names(dat)
str(dat)
```

```{r fig.width=12, fig.height=10}
dat = dat %>% drop_na()

ggplot(data=dat, aes(Year, fill=Gender)) + geom_bar(position='dodge') + ggtitle("Respondents") + theme(plot.title = element_text(hjust = 0.5), axis.title.y=element_blank()) + xlab("Year") + labs(fill="Gender")
```

```{r fig.width=12, fig.height=10}
ggplot(data=dat, aes(Year, fill=PolParty)) + geom_histogram(position='dodge', binwidth = 10) + theme(legend.position="right") + ggtitle("Astrological Sign of the Respondent") + theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank(), axis.title.x = element_blank(),axis.text.x=element_blank(), axis.ticks.x=element_blank() )
```


```{r fig.width=12, fig.height=10}
ggplot(data=dat, aes(PolParty, fill=PolParty)) + geom_bar() + theme(legend.position="right") + ggtitle("Happiness") + theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank(), axis.title.x = element_blank(),axis.text.x=element_blank(), axis.ticks.x=element_blank() ) + facet_wrap((~Happy))
```



```{r fig.width=12, fig.height=10}
ggplot(data=dat, aes(Sign, fill=Sign)) + geom_bar() + theme(legend.position="right") + ggtitle("Astrological Sign") + theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank(), axis.title.x = element_blank(),axis.text.x=element_blank(), axis.ticks.x=element_blank() ) + facet_wrap((~Happy))
```



```{r fig.width=12, fig.height=10}
ggplot(data=dat, aes(Sign, fill=Sign)) + geom_bar() + theme(legend.position="right") + ggtitle("Astrological Sign") + theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank(), axis.title.x = element_blank(),axis.text.x=element_blank(), axis.ticks.x=element_blank() ) + facet_wrap((~PolParty))
```
```{r fig.width=12, fig.height=10}
ggplot(data=dat, aes(School, fill=School)) + geom_bar() + theme(legend.position="right") + ggtitle("Country of Origin") + theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank(), axis.title.x = element_blank(),axis.text.x=element_blank(), axis.ticks.x=element_blank() ) + facet_wrap((~Origin))

```



```{r fig.width=12, fig.height=10}
ggplot(data=dat, aes(Happy, fill=Happy)) + geom_bar() + theme(legend.position="right") + ggtitle("Country of Origin") + theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank(), axis.title.x = element_blank(),axis.text.x=element_blank(), axis.ticks.x=element_blank() ) + facet_wrap((~Origin))

```



```{r fig.width=12, fig.height=10}
ggplot(data=dat, aes(PolParty, fill=PolParty)) + geom_bar() + theme(legend.position="right") + ggtitle("Country of Origin") + theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank(), axis.title.x = element_blank(),axis.text.x=element_blank(), axis.ticks.x=element_blank() ) + facet_wrap((~Origin))

```



```{r fig.width=12, fig.height=10}
ggplot(data=dat, aes(Happy, fill=Happy)) + geom_bar() + theme(legend.position="right") + ggtitle("Education") + theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank(), axis.title.x = element_blank(),axis.text.x=element_blank(), axis.ticks.x=element_blank() ) + facet_wrap((~School))
```
1.	Find the average years of education for the respondents with each political party affiliation. Arrange your output in a meaningful order. Describe any pattern you find.
```{r}
dat <- dat %>%
  mutate(School = as.numeric(School))
str(dat)
dat %>%
  group_by(PolParty) %>%
  summarize(AvgEdu = mean(School, na.rm = TRUE), MedEdu = median(School, na.rm = TRUE)) %>%
  arrange(AvgEdu)

```
2.	Create a single data frame containing records for the better educated respondents in each political party affiliation. A “better educated” respondent is someone who has more years of education than the median within the same party affiliation. Print the structure of the data frame.
```{r}
PolPartyEdu <- dat %>%
  select(PolParty, School)

PolPartyEdu <- PolPartyEdu %>%
  group_by(PolParty) %>%
  filter(School >= median(School, na.rm = TRUE))

str(PolPartyEdu)

```

3.	How is the happiness of a respondent related to his/her marriage status?
```{r fig.width=12, fig.height=10}
ggplot(data=dat, aes(Married, fill=Married)) + geom_bar() + theme(legend.position="right") + ggtitle("Married") + theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank(), axis.title.x = element_blank(),axis.text.x=element_blank(), axis.ticks.x=element_blank() ) + facet_wrap((~Happy))
```

4.	Does party affiliation affect the relationship you found in the last question?
```{r fig.width=12, fig.height=10}
ggplot(data=dat, aes(x=Married, fill=PolParty)) + geom_bar() + theme(legend.position="right") + ggtitle("Married") + theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank(), axis.title.x = element_blank(),axis.text.x=element_blank(), axis.ticks.x=element_blank()) + facet_wrap((~Happy))
```

5.	How is marital status related to the education of a respondent? What are possible reasons accounting for the patterns you see?

Interestingly, there really isn't a big swing in the range of martial status, even the median is similar across the groups (except for *No answer*).   

```{r fig.width=12, fig.height=10}
ggplot(data=dat, aes(x=Married, y=School, fill=Married)) + geom_boxplot() + theme(legend.position="right") + ggtitle("Marital Status vs Education") + theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank(), axis.title.x = element_blank())
```

6.	Explore two more interesting questions, and answer those questions using the GSS data.

```{r}
dat <- dat %>%
  mutate(Vocab = as.numeric(Vocab))
##str(dat)

YearEd <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, "Don't know", "No answer")

ggplot(aes(x=School, y=Vocab, group=School), data=dat) + geom_boxplot() + ggtitle("Education of Vocabulary") + theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ylab("Vocabulary") + xlab("Years of Education") + scale_x_discrete(limits = YearEd) + scale_x_continuous(limits = c(-1,21))
```

```{r}
dat <- dat %>%
  mutate(Vocab = as.numeric(Vocab))
##str(dat)

##ggplot(dat, aes(School, fill=School)) + geom_bar() + facet_wrap(~Sign)

ggplot(data=dat, aes(Sign, fill=Sign)) + geom_bar() + theme(legend.position="right") + ggtitle("Astrological Sign") + theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank(), axis.title.x = element_blank(),axis.text.x=element_blank(), axis.ticks.x=element_blank() ) + facet_wrap((~School))

##ggplot(aes(School, fill=School), data=dat) + geom_bar() + ggtitle("Education of Vocabulary") + theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ylab("") + xlab("Years of Education") + facet_wrap(~Sign)
```