---
title: "Lab 3 - Team 5"
author: "Lisa McDaniel"
date: "9/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
**Lisa McDaniel - Exploratory Analysis Summery**




**John King - Exploratory Analysis Summary**




```{r}
library(readxl)
library(tidyverse)
dat <- readxl::read_xls('../ds202lab3/GSS.xls', sheet="Data")
names(dat)
str(dat)
dat
```

```{r echo=FALSE}
dat <- dat %>%
  rename(Sign = "Respondents astrological sign" ) %>%
  rename(Vocab = "Number words correct in vocabulary test") %>%
  rename(Origin = "Country of family origin") %>%
  rename(MarHap = "Married people happier than unmarried") %>%
  rename(Happy = "General happiness") %>%
  rename(PolParty = "Political party affiliation") %>%
  rename(Gender = "Respondents sex") %>%
  rename(School = "Highest year of school completed") %>%
  rename(Occ80 = "Rs occupational prestige score  (1980)") %>%
  rename(Occ70 = "Rs occupational prestige score  (1970)") %>%
  rename(IDnum = "Respondent id number") %>%
  rename(Ballot = "Ballot used for interview") %>%
  rename(Year = "Gss year for this respondent") %>%
  rename(Married = "Marital status")

names(dat)
str(dat)
```
 1. Find the average years of education for the respondents with each political      party affiliation. Arrange your output in a meaningful order. Describe any       pattern you find.
  
  **Observation**` 
```{r}
dat = dat %>% drop_na()
dat <- dat %>%
  mutate(School = as.numeric(School))

dat %>%
  group_by(PolParty) %>%
  summarize(AvgEdu = mean(School, na.rm = TRUE), MedEdu = median(School, na.rm = TRUE)) %>%
  arrange(AvgEdu)

```


  2. Create a single data frame containing records for the better educated            respondents in each political party affiliation. A “better educated”             respondent is someone who has more years of education than the median within      the same party affiliation. Print the structure of the data frame.
 
```{r}
PolPartyEdu <- dat %>%
  select(PolParty, School)

PolPartyEdu <- PolPartyEdu %>%
  group_by(PolParty) %>%
  filter(School >= median(School, na.rm = TRUE))

str(PolPartyEdu)

```  
  

  3. How is the happiness of a respondent related to his/her marriage status?
  
  
  
  4. Does party affiliation affect the relationship you found in the last             question?
  
  
  
  5. How is marrital status related to the education of a respondent? What are        possible reasons accounting for the patterns you see?
  
  
  
  6. Explore two more interesting questions, and answer those questions using the      GSS data.
